---
title: 'Experiment 3: More airline data'
author: "Cole Bennett"
date: '2023-09-11'
output: html_document
---

This experiment (conceptually) continues our class work on flight data.
We will continue to refer to (https://r4ds.hadley.nz/data-transform).

** However, we will expand and update our dataset by going to a root source ---
the Bureau of Transportation.

Go to website:
https://www.transtats.bts.gov/DL_SelectFields.aspx?gnoyr_VQ=FIM&QO_fu146_anzr=Nv4%20Pn44vr45

This dataset allows access to information on domestic flight and domestic carriers, with monthly summary data.

# packages
```{r}
library(nycflights13)
library(tidyverse)
library(janitor)
library(here)
```

```{r}
thisfile = here("data_raw", "DL_SelectFields.kmz")

df2022 = read_csv(thisfile) |> clean_names()
```
#Clean Names is from the Janitor package, and it lets you transform everything into a unanimous "case". In this set, some of it was in Snake Case, but the capitalization was inconsistent, so we want to "clean_names()"

# This is a LOT of information, so let's pick some data of interest and Subset it...
```{r}
df9 = df2022 |>
  filter(passengers > 0, origin == "LGA")
```
# This is only flights from Laguardia and that actually are carrying passengers and not just cargo.


#Let's try to visualize a monthly count of Flights using a barchart.

```{r}
df9 |>
  ggplot(aes(month)) +
  geom_bar()
```
# This is kind of a flawed graph though...
# By default, geom_bar() is counting the number of rows, where we have asked it to visualize the count by "month". (The airlines are reporting all numbers by month)
## ...If you look at the data set, you'll see why counting rows isn't going to give us a count of flights...

### This is what we want:
# (Weight each row by the number of departures actually performed)
```{r}
df9 |>
  ggplot(aes(month)) +
  geom_bar(aes(weight = departures_performed))
```

### A new visualization
# Make a boxplot where the bar height is based on the number of passengers riding each month?
```{r}
df9 |>
  ggplot(aes(month)) +
  geom_bar(aes(weight = passengers))
```
### Observation:
# Low passenger numbers in early 2022, probably due to COVID.

## Just because we can...

# This is a pretty plot, but not a useful plot...
```{r}
df9 |>
  ggplot(aes(month, fill = carrier_name)) +
  geom_bar(aes(weight = departures_performed))
```

### Now lets make a data frame of Arrivals AND Departures from Laguardia.
```{r}
df10 = df2022 |>
  filter(passengers > 0, origin == "LGA" | dest == "LGA")

df10 |> ggplot(aes(month)) +
  geom_bar(aes(weight = passengers))
```
# remeber that | is read as "or".

## We don't care about ALL the columns of variables. Just a few.
```{r}
df11 = df10 |>
  select(month, passengers, seats, carrier_name, dest, origin)
```

# can also do it this way:
```{r}
df12 = df10 |>
  select(1:5, month, contains("id"))
```
# the 1:5 just selects the first 5 variable coulmns.

### How Full are the Airplanes on average???
```{r}
df13 = df11 |>
  mutate(percent_loading = passengers / seats * 100)

df13 |> ggplot(aes(percent_loading)) +
  geom_histogram() +
  facet_wrap(~carrier_name, scales = "free_y")
```

