---
title: "experiment 1: airline data analysis"
date: "2023-09-06"
output: html_notebook
---
# packages we're using for this analysis
```{r}
library(tidyverse)
library(here)
library(janitor)
library(nycflights13)
```

# data
# creating the "base" data-frame we are interested in...
```{r}
flights |>
  filter(dest == "IAH") |> 
  group_by(year, month, day) |> 
  summarize(
    arr_delay = mean(arr_delay, na.rm = TRUE)
  )

df1 <-  flights

```
# we now see a data frame that depicts the mean delay time of all flights going into "IAH". One row for each day of the year


# flights from Miami in Sept.
```{r}
df2 <-  df1 |> filter(origin == "MIA", month == 9)

```

# flights going to Miami in Sept.
```{r}
df3 <- df1 |> filter(dest == "MIA", month == 9)

glimpse(df3)
```
# flights going to Miami in Jan.
```{r}
df4 <- df1 |> filter(dest == "MIA", month == "1")

```
# flights going to Chicago during the summer.
```{r}
df5 <- df1|> filter(dest == "MDW", month == (6 | 7 | 8))

```
# unique flight numbers of flights going to Miami
```{r}
distinctdf3 = df3 |> distinct(flight)

view(distinctdf3)
```
### After viewing "distinctdf3", it's easy to identify the smallest flight number, which is 83. You can also literally just view df3 and click the tab to sort dest from smallest to biggest. 

# All the flights in 2013 going to Miami, that are of the smallest flight number (83)...
```{r}
df6 = df1 |> filter(flight == 83, dest == "MIA")

view(df6)
```

# (Plot) Observing if there is a trend between departure delay and arrival delay, for Flight 83.
```{r}
ggplot(data = df6,
       mapping = aes(x = dep_delay, y = arr_delay)) + 
  geom_point()
```

# Histogram showing frequencies of the overall delays (depature delay minus the arrival delay) of Flight 83.
# This reveals a skew to the left...
```{r}
ggplot(data = df6,
      mapping = aes(x = dep_delay-arr_delay)) + 
  geom_histogram(binwidth = 10)
```


### Questions to answer for Homework (9/11/23):

#1 Is flight time affected by delayed departure. (Do the airlines try to "catch up"?)
#2 Does the departure delay change across time of day (later in the day has more delays.)
#3 Is flight time pattern affected by time of year
#4 Is departure delay affected by time of year.

```{r}
df1 |>
  filter(dest == "MIA") |>
  count(origin, sort = TRUE)
```
# count shows all the unique observations under origin, and then shows the frequency of that observation. 
# sort = TRUE makes it so that it's ordered from most frequent to least. (idk why that does that but thats what it does)

### Is flight time affected by delayed departure?
#I want to examine whether the flight time is impacted by delayed departure.
#I want to compare flight time to 'planned' flight time. So we create a new variable "flt_delta". 
__Mutate: Is how we make new Variables!__
#A flight that arrives 10 mins late, if it departed on time, had a "delta" of 10 mins over.
```{r}
df7 <- df1 |>
  filter(dest == "MIA", origin == "LGA") |>
  mutate(flt_delta = arr_delay - dep_delay)
```

```{r}
df7 |> 
  ggplot(aes(x = dep_delay, y = flt_delta)) +geom_point(alpha = .1)
```
# alpha sets a "transperancy" of the dots. this way we can visualy see the concentration of dots via areas with the darker shades.

# We can see from this that Flight 83 tries to always be early, regardless of departure delay.

```{r}
df7 |> 
  ggplot(aes(x = dep_delay, y = flt_delta)) +geom_point(alpha = .1) +geom_hline(aes(yintercept = mean(flt_delta, na.rm = TRUE)))
```
# You can see that with any flights with a depature delay, they do their best to make it up.

### Is Departure Delay affected by Time of Year?
```{r}
df7 |> 
  ggplot(aes(x = time_hour, y = dep_delay)) +geom_point(alpha = .1)
```

```{r}
df7 |> 
  ggplot(aes(x = time_hour, y = dep_delay)) + 
  geom_point(alpha = .1) +
  stat_smooth() +
  ylim(-25,120)
```
# We can see that throughout the year, the flights actually tend to leave early if they can. Most of the data is under y = 120 (dep_delay)
### Why are delays bigger in December than in January? Probably not the weather... ###

### Does the departure delay change across time of day?
```{r}
df7 |>
  ggplot(aes(x = hour + minute / 60, y = dep_delay)) +
  geom_point(alpha = .1) +
  stat_smooth() +
  ylim(-25,120)
```
# This is fascinating, because it shows that later in the day, there are more departure delays. Departure delay increases across the flight day.

### What about by day of the week?

```{r}
df7 |>
  mutate(day_of_week = weekdays(time_hour)) |>
  ggplot(aes(x = hour + minute / 60, y = dep_delay, color = day_of_week)) +
  geom_point(alpha = .1) +
  stat_smooth() +
  ylim(-25,120)
```
# The days are sorted alphabetically by default. We'll learn how to change this later.
# Thursday seems to have the worst delays. It's difficult to view them all on top of each other, so lets separate them all out individually by "faceting" the data.

```{r}
df7 |>
  mutate(day_of_week = weekdays(time_hour)) |>
  ggplot(aes(x = hour + minute / 60, y = dep_delay, color = day_of_week)) +
  geom_point(alpha = .1) +
  stat_smooth() +
  ylim(-20,40) +
  facet_wrap(~day_of_week)
```
# You can definetly observe that Thursday is the worst day to fly out of Laguardia...